Stage-Splitting a Modal Language

### 2014-09-24 paper/semantics.tex
dumped some sentences for a new section 2

\section{The \lang\ Language}
\label{sec:semantics}

\input {figures/grammars}
\input {figures/statics}

In this section, we describe \lang, a simple two-stage language
for which we define and analyze stage splitting.

% \lang\ is largely based off the
% language \lamCircle\ by \cite{davies96}.  
% which was used as a model for partial evaluation.  
% Our version, despite a slightly different semantics, can be used for the same purpose.
% The details of the partial evaluation view are shown in \cref{sec:partialeval}.

\lang\ is essentially two copies of a typed lambda calculus with
products, sums, and isorecursive types.%
\footnote{It is possible for different stages to have different sets of
features, but for simplicity we do not consider this.}
Every valid expression has both a type and a stage, either \bbone~or \bbtwo,
expressing in which copy of the language the term resides. Intuitively, the
stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.

>>>>

\section {The \lang\ Language}

We now describe \lang, a simple two-stage language
for which we will later define and analyze a stage splitter.

For this purpose, we use {\em fast exponent} as a running example.
It calculates $b^p$ in $O(\log p)$ time.  
Using a standard unstaged ML-like syntax, it is
\begin{lstlisting} 
fun fexp (b : int, p : int) : int = 
	if p == 0 then
		1
	else if (p mod 2) == 0 then
		let x = fexp(b,p/2) in x*x
	else
		b * fexp(b,p-1)	
\end{lstlisting}

The main idea of \lang\ is that terms defined in it are 
split across two {\em stages}, called \bbone\ and \bbtwo,
such that information can flow from the stage \bbone\ portions of the term to the stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.
Our starting point for this is a typed lambda calculus with products, sums, and isorecursive types.
Here is the fast exponent example in staged form:
\begin{lstlisting} 
let fexp (b : $int, p : int) : $int =
	if p == 0 then
		next{1}
	else if (p mod 2) == 0 then
		next{let x = prev{fexp(b,p/2)} in x*x}
	else
		next{prev{b} * prev{fexp(b,p-1)}}		
\end{lstlisting}
This staged version contains three new constructs: 
$\next$ and $\prev$ at the term level, and $\fut$ at the type level.
We can read $\next$ and $\prev$ as indications to 
execute the code in the following block at stage \bbtwo\ and \bbone, respectively.
The type constructor $\fut$ is a way for us to talk at stage \bbone\ about things that will not be available until stage \bbtwo.
Thus, ${\tt fexp} : \fut\rmint \times \rmint \to \fut\rmint$ 
can be read as a function that takes a base at stage \bbtwo\ and a power at stage \bbone, 
and returns the exponent at stage \bbtwo.

With this in mind, observe that the exponent decomposition and recursive calls in ${\tt fexp}$ 
all occur in stage \bbone, whereas squaring and multiplication by the base occur at stage \bbtwo.

This paper investigates a transformation called stage-splitting, 
wherein a single multi-stage program is split into multiple single-stage programs.
For ${\tt fexp}$, this is:

\begin{lstlisting} 
fun fexp1 (b, e) =
  ((), roll (
    if e == 0 then inj1 ()
    else
      inj2 (
        if (e mod 2) == 0
        then inj1 (#2 (fexp1 (b,e/2)))
        else inj2 (#2 (fexp1 (b,e-1)))
      )
  ))

fun fexp2 ((b, e), p) =
	case unroll p of
	  inj1 () => 1
	| inj2 d =>
		case d of
		  inj1 r => let x = fexp2 ((b,()),r) in x*x
		| inj2 r => b * fexp2 ((b,()),r)
\end{lstlisting}

\section{Statics and Dynamics}
\label{sec:semantics}

\input {figures/grammars}
\input {figures/statics}
%
\footnote{It is possible for different stages to have different sets of
features, but for simplicity we do not consider this.}
Every valid expression has both a type and a stage, either \bbone~or \bbtwo,
expressing in which copy of the language the term resides. Intuitively, the
stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.
